Syntax Is All You Need: A Universal-Language Approach to Mutant Generation

### 2023-09-27 paper/main.tex
fits

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  While mutation testing has been a topic of academic interest for
decades, it is only recently that ``real-world'' developers, including
industry leaders such as Google and Meta, have adopted mutation
testing.  In this paper we propose a new approach to the development of mutation
testing tools, and in particular the core challenge of
\emph{generating mutants}.  Current practice tends towards two
limited approaches to mutation generation: mutants are either (1)
generated at the bytecode/IR level, and thus neither human readable
nor adaptable to source-level features of languages or projects, or
(2) generated at the source level by language-specific tools that are
hard to write and maintain, and in fact are often abandoned by both
developers and users.  We propose instead that source-level mutation
generation is a special case of \emph{program transformation} in
general, and that adopting this approach allows for a single tool that
can effectively generate source-level mutants for essentially
\emph{any} programming language. Furthermore, by using \emph{parser
  parser combinators} many of the seeming limitations of an
any-language approach can be overcome, without the need to parse
specific languages.  We compare the universal program transformation
approach to mutation to existing tools, and demonstrate the advantages
of using parser parser combinators to improve on a simple regular-expression
based approach to generation.  Finally, we show that our approach
can provide effective mutant generation even for a language for which
it lacks any language-specific operators, and that is not very similar
in syntax to any language it has been applied to previously.
\end{abstract}

@@

In this paper, we argue that developing a large number of mutation
testing tools, each requiring its own development community, workflow
for mutation, and effort to keep up with language changes, is fundamentally
unnecessary.  The core problem of mutation testing,
\emph{mutant generation}, is simply a particular instance of the
problem of \emph{program transformation}.  Program transformation
considers the problem of taking a program, $P$, and producing a
program, $P'$ that is the result of applying some function $f$ to $P$
(i.e., $P' = f(P)$).  In mutant generation, we
consider a family of functions $f$ where each $f_{o,loc}$ is an
instantiation of a \emph{mutation operator} $o$  (as classically, if
often informally, defined in the literature) at a particular program
location $loc$ in $P$.  The inclusion of $loc$ in the specification of
$f$ indicates that the problem is a specialization of general
program transformation in that all classic mutation operators we are
aware of are \emph{local}: they only affect a small portion of a
program, genearlly a contiguous string of characters or tokens.

@@

      \item A comparison of an implementation of mutant generation
       based on this approach with existing tools for four important
       programming languages.  The single-language tools range from
       approximately 3,800 LOC to nearly 60,000 LOC, and, we emphasize, each handle
       one programming language.  Our tool, which provides comparable
       core mutation testing functionality, supports over a dozen
       languages using only about 2,200 LOC of Python and less than 500
       lines for rules defining mutation operators over a dozen
       specific languages (and which can effectively mutate an
       essentially unlimited number of other languages).  Our
       approach, we emphasize, enables a smaller tool than any other tool
       we examined (by over 1000 LOC) to support a rich set of
       operators (the full set used by Google's internal mutation
       tools, plus a large number of additional operators) for all languages covered by other tools, plus a set
       of additional languages.
       \item Far from being limited by its implementation size and
         universality, our tool \emph{generates a larger number of valid,
         non-equivalent mutants} for every language we examined than
         any competing, language-specific tool.  That these additional
         mutants
         are meaningful, rather than obscure, is further supported by
         the fact that our mutation scores are similar to those for
         other tools.

@@

In Python regular expression syntax, this expresses the replacement of
strings with a possibly-empty amount of white space followed by a
character pattern indicating a line of source code, with the same
source code, preserving indentation, turned into a comment.
While much more expressive than simple string replacement, regular
expressions are familliar to most developers.  In principle,
development of complex regular expressions for operators can even be
performed by simply providing examples~\cite{bartoli2014automatic}  or
describing the operator in natural
language~\cite{zhong2018generating}.  
These approaches speak to the potential of string manipulation-based approaches
for mutation testing, without the need for program parsing or analysis.

@@

\begin{table}[hbtp]
\centering
\small 
\caption{Regex vs. Comby for Python. LOC describes the project files
  considered. Muts. is total number of mutants generated for those
  files.  Invalid mutants are ones that fail module load (e.g., due
  to imbalanced parenthesis, bad indentation, or similar problems).  We can use TCE to check for redundant Python mutants, shown under ``Red.''}
\label{tab:table_python1}
\begin{varwidth}{.9\columnwidth}
\begin{tabularx}{\linewidth}{l|r|rcrc|rcrc}
\toprule
\toprule
                 &                 & \multicolumn{4}{c|}{\textbf{Regex}}             &\multicolumn{4}{c}{\textbf{Comby}}  \\[1ex]
\textbf{Project} & \textbf{LOC}    & \textbf{Muts.} & \textbf{Red.}  & \textbf{\% Valid} & \makecell{\textbf{Mut.} \\ \textbf{Score}} & \textbf{Muts.} & \textbf{Red.}   &  \textbf{\% Valid} & \makecell{\textbf{Mut.} \\ \textbf{Score}}\\\midrule
keract           & 402             & 3984  & 729   & 41.53\% & 0.28           & 3067  & 533    & 47.96\%    & 0.26\\
dtw              & 118             & 2106  & 567   & 46.20\%    & 0.52           & 1835  & 639    & 47.19\%  & 0.55 \\
notion-          & \mr{2}{189}     & \mr{2}{1038}  & \mr{2}{82}    & \mr{2}{18.89\%}   & \mr{2}{0.57}         & \mr{2}{1019}  & \mr{2}{122}    & \mr{2}{19.23\%} & \mr{2}{0.60}  \\
sdk-py           & & & & & & \\[0.5ex]
atlassian-       & \mr{2}{236}      & \mr{2}{1511}  & \mr{2}{207}   & \mr{2}{19.99\%}    & \mr{2}{0.64}        & \mr{2}{1319}  & \mr{2}{351}    & \mr{2}{25.25\%} & \mr{2}{0.62}  \\
python-api       & & & & & & \\[0.5ex]
ESD              & 698             & 6268  & 1642  & 29.46\%    & 0.07           & 3874  & 614    & 44.18\% & 0.07\\
fiber            & 358             & 1565  & 163   & 31.08\% & 0.62           & 1466  & 241    & 31.24\% & 0.64\\\bottomrule
\end{tabularx}
\begin{tabularx}{\linewidth}{Xllrrrrc}
                  && &  \textbf{Red.}     & \textbf{Valid}  & \textbf{Invalid}  & \textbf{\% Valid} &\textbf{Mut. Score}\\\midrule
\multirow{4}{*}{\textbf{Overall}} & \multirow{2}{*}{Regex} & Mean &  565.00 & 913.00 & 1267.33 & 31.19\%  & 0.45\\
    &   &  Std.Dev. & 584.79  & 708.93 & 813.35 & 11.06\%  & 0.23\\\cline{2-8}
 & \multirow{2}{*}{Comby} & Mean & 416.67 & 839.33 & 840.67 & 35.84\% & 0.46\\
  &   &  Std.Dev. &  211.62 & 628.79 & 418.99 & 12.28\%  & 0.24 \\\bottomrule
\end{tabularx}
\end{varwidth}
\end{table}

@@

We do not report the time taken for mutation generation and testing.
Dextools uses mutant schema, and PIT performs bytecode mutation, so
both are much faster for compiling mutants, but other tools took similar
amounts of time to compile mutants.   Mutant execution time was also
generally similar, except for tools that produced very few mutants.
We expect that our approach is generally slower for initial mutant
generation, due to a higher number of invalid mutants, but that this
cost is relatively small compared to the time required to compile and
execute valid mutants for almost all projects.  An informal study of
large Java projects by the author of the PIT gradle
plugin\footnote{\url{https://solidsoft.wordpress.com/2017/09/19/how-fast-or-slow-mutation-testing-really-is/}}
  suggests that even a successful compile of a typical project takes
  somewhere between one-sixth to one-third the time required for
  executing tests, and failed compiles are usually faster than
  successful ones, since invalid mutants almost always fail during
  parsing, and so code generation, optimization, and linking are not performed.

@@

The ease of defining mutation operators,
plus our interest in generating large numbers of mutants and
prioritizing the most interesting, in order to make mutation testing useful for
projects with extremely high quality test suites, or even using
projects using formal
verification (as in the work of Groce et al.~\cite{groce2018verified}), has encouraged us to take a ``let a thousand flowers bloom''
approach to choosing mutation operators.  For example, our approach
provides many more control-flow mutations than most tools, due to our
observation that test suites sometimes fail to check for loop
execution counts other than one and zero.  Few other tools introduce
{\tt break} and {\tt continue} statements into programs, or transform
one into the other.  These mutants are often equivalent (in fact, they
give rise to a large number of our equivalent mutants reported above), but \emph{also
sometimes expose subtle weaknesses in a high quality test suite}, in our
experience.  In general, we take our philosophy of mutation from the
proposal by Gopinath et al.~\cite{Limits,gopinath2017mutation} that
expanding the set of mutation operations and classes of such
operations, in the long run, is more important than using only a small
set of ``good'' operators, in that the end-goal of mutation testing is
to identify weaknesses in test suites, not simply to estimate a single
score.   The price paid for this
expansiveness in allowing ways to detect weaknessess in testing is a
larger number of mutants to execute, and a slightly larger number of
equivalent mutants.  However, this is not a
fundamental limitation or forced choice.  We propose as future work to
split the operators for each level of the hierarchy into (at least) two sets, one
set of core operators similar to (but larger than) those provided by
LittleDarwin or Cargo-Mutants, and a set of ``aggressive'' operators
that includes our full current set of operators, a very simple matter due to
our approach to mutant generation.

In fact, this brings us back to a point made in the introduction.
While our approach introduces some unavoidable costs in the
(relatively cheap, compared to execution and analysis) compilation stage of
mutation testing, it may offer \emph{more} flexibility in addressing
execution and analysis stages.  Defining limited custom sets of
operators to run is trivial in our setting, making it easy to
execute only, e.g., statement mutations, or, in a numeric-computation
intensive codebase, only arithmetic operators.  Expanding the
hierarchical structure to include ``famillies'' of mutation operators
is trivial, and lets users specify fine-grained mutation
generation for all languages used in a project, in a simple, consistent way.  Our approach also produces actual source
copies of all generated mutants, making it easy to apply techniques
like Predictive Mutation Testing~\cite{zhangPMT} that work with source
code.  This also makes it easy to integrate, as we already have done, heuristic
aids to mutant \emph{analysis} such as ranking of unkilled mutants by
similarity and ``interstingness,'' an approach that has already proven
useful in using mutation testing to improve static analysis tools~\cite{StatMut}.
\subsection{Haskell Case Study}

@@

These mutants produce a mutation
score of 0.8 for the two properties defined in the original MuCheck
paper.  In other words, like MuCheck, our tool is able to
determine that the properties provided are insufficient.  In
particular, while they check idempotence and that the result is
sorted, they do not check that the result of {\tt qsort} is a permutaton of the input
list.  The MuCheck paper only provides mutation scores for the two
properties independently.  For idempotence, MuCheck gives a mutation
score of 0.84, and our implementation yields a score of 0.8.  Checking
only sortedness, MuCheck yields a score of 0.61, while our approach
produces an even lower score of 0.4.  Our lower
scores are not due to equivalent mutants: adding a property to check
that the sorted list is a permutation of the original list kills all
our mutants.

Of course, our implementation does not provide the small set of Haskell-specific
mutations provided by MuCheck (e.g., type-aware function replacement).  It nonetheless produces clearly useful
mutants, even for a language arguably radically different from those
for which the universal rules were developed.  In short, our approach
likely provides users of a new language with mutation testing,
\emph{even if no effort has been expended to define rules for the
  language at all}.

@@

\begin{lstlisting}[language=haskell,caption={\textbf{GPT-4 Generated Haskell Mutation Tool}},label=lst:mutationTool]
-- Perform mutation on a module
mutateModule :: Module -> Module

-- Perform mutation on an expression
mutateExp :: Exp -> Exp
mutateExp (If cond trueBranch falseBranch) = If cond falseBranch trueBranch
mutateExp x = x
...
\end{lstlisting}

@@

\begin{lstlisting}[caption={\textbf{GPT-4 Mutants on a simple Haskell Program}},label=lst:haskellmutants]
[0..] ==> [1..]
IO () ==> IO String  
"Current TODO list:" ==> "Updated TODO list:"
"Invalid command: " ==> "Unrecognized command: "
\end{lstlisting}

To understand how our approach compares with state of the art large language models (LLM), we prompt GPT-4 to generate a mutation testing tool for Haskell.
Surprisingly, the generated program in \cref{lst:mutationTool} has no syntax errors, and even runs to produce a mutant. However, the program only implements 
the inverting conditionals operation and some basic module mutations,
a small subset of all the operators our tool implements, despite our
tool being unaware of Haskell's existence. Furthermore, like existing mutation testing tools for source code,
the generated code performs AST manipulation in Haskell, making it harder to read than the Comby rules of our approach. Asking the LLM to expand the set of rules
 produces more operators; however the number of mutants generated by
 such an approach would be a small subset of what our approach is
 capable of doing, even for a language we do not technically support.

 Another approach is to have LLMs generate mutants for a given program. We experiment with prompting GPT-4 to generate mutants for a simple Haskell Todo-list management program.
 \cref{lst:haskellmutants} shows the mutants that GPT-4 generates for this program (formatted as before ==> after). Notice that some of the mutants generated make sense, and 
 resemble bugs a developer might make (replacing 0 with 1 or using the wrong data type). However, GPT-4 also generates equivalent mutants too, generating 
 (essentially) equivalent strings for all output to console. One is unlikely to generate tests that can kill these mutants, unless checking the exact string contents of all 
 logging messages. The error messages (potentially) could be checked,
 but the semantically equivalent meaning of these strings makes them
 poor mutants (which is why no mutation tools we are aware of
 implement such mutants).

While LLMs are not able to generate fully working language specific mutation testing tools from scratch, they are capable of generating many interesting rules
 that are specific to a given language. For example, prompting GPT-4
 to generate Haskell rules for our tool generates the set of rules seen in \cref{lst:haskellcomby}. The generated rules are 
 both valid Comby syntax and language-specific. Specifically, replacing head with tail is a rule that would only apply to functional programming languages such as 
 Haskell.  This suggests that while extending other tools to include new rules
 often involves writing complex code to manipulate ASTs, our tool may
 be extensible simply by requesting an LLM to implement a natural
 language description of the desired rule, since GPT-4 already knows
 how to use Comby (and write regular expressions).

@@

Hariri et al. compared C mutation approaches at the source and
compiler IR levels~\cite{CompareSrcBinary} and found that overall
source level mutation was better, producing fewer mutants overall, but
matching the IR approach in the important measures of surface and
minimal mutants and overall mutation score.  Numerous studies compare
Java mutation tools~\cite{MajorPIT,gopinath2017does}, including a
recent article for a more general audience in Communications of the
ACM~\cite{CommACMJavaTool} (perhaps showing the growing interest in
practical mutation testing).  This paper showed that users
considered active maintenance, support for a variety of testing
frameworks, and support for recent Java versions as the most important
features in a Java mutation tool.  The approach proposed in this paper
by its nature tends to promise all three of these key factors without
imposing burdens on maintainers. More recently, there has been some effort to create a 
language agnostic mutation testing tool~\cite{wodeltest}\footnote{An
  additional paper on this general topic is a 4 page tool paper for an initial
  prototype of our current tool, omitted for blinding.}. However, this tool uses
Wodel, a new (and hard-to-read, compared to Comby) DSL for expressing operators, and requires users to
define a set of operators for each language, without sharing
of rules.

@@

As future work, we plan to give users more control over which mutation
operators are used, with controls common to multiple languages.  We
add that our approach in general seems to be a good fit to modern
large software projects, which are often
multilingual~\cite{Multilingual}: a single mutation tool can be used
for an entire project.

>>>>

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  While mutation testing has been a topic of academic interest for
decades, it is only recently that ``real-world'' developers, including
industry leaders such as Google and Meta, have adopted mutation
testing.  We propose a new approach to the development of mutation
testing tools, and in particular the core challenge of
\emph{generating mutants}.  Current practice tends towards two
limited approaches to mutation generation: mutants are either (1)
generated at the bytecode/IR level, and thus neither human readable
nor adaptable to source-level features of languages or projects, or
(2) generated at the source level by language-specific tools that are
hard to write and maintain, and in fact are often abandoned by both
developers and users.  We propose instead that source-level mutation
generation is a special case of \emph{program transformation} in
general, and that adopting this approach allows for a single tool that
can effectively generate source-level mutants for essentially
\emph{any} programming language. Furthermore, by using \emph{parser
  parser combinators} many of the seeming limitations of an
any-language approach can be overcome, without the need to parse
specific languages.  We compare this new
approach to mutation to existing tools, and demonstrate the advantages
of using parser parser combinators to improve on a regular-expression
based approach to generation.  Finally, we show that our approach
can provide effective mutant generation even for a language for which
it lacks any language-specific operators, and that is not very similar
in syntax to any language it has been applied to previously.
\end{abstract}

@@

Developing a large number of mutation
testing tools, each requiring its own development community, workflow
for mutation, and effort to keep up with language changes, may, however, be
unnecessary.  The core problem of mutation testing,
\emph{mutant generation}, is simply a particular instance of the
problem of \emph{program transformation}.  Program transformation
considers the problem of taking a program, $P$, and producing a
program, $P'$ that is the result of applying some function $f$ to $P$
(i.e., $P' = f(P)$).  In mutant generation, we
consider a family of functions $f$ where each $f_{o,loc}$ is an
instantiation of a \emph{mutation operator} $o$  (as classically, if
often informally, defined in the literature) at a particular program
location $loc$ in $P$.  The inclusion of $loc$ in the specification of
$f$ indicates that the problem is a specialization of general
program transformation in that all classic mutation operators we are
aware of are \emph{local}: they only affect a small portion of a
program, genearlly a contiguous string of characters or tokens.

@@

      \item A comparison of an implementation of mutant generation
       based on this approach with existing tools for four important
       programming languages.  The single-language tools range from
       approximately 3,800 LOC to nearly 60,000 LOC, and, we emphasize, each handle
       one programming language.  Our tool, which provides comparable
       core mutation testing functionality, supports over a dozen
       languages using only about 2,200 LOC of Python and less than 500
       lines for rules defining mutation operators over a dozen
       specific languages (and which can effectively mutate an
       essentially unlimited number of other languages).  Our
       approach enables a smaller tool than any other tool
       we examined (by over 1000 LOC) to support a rich set of
       operators for all languages covered by other tools, plus additional languages.
       \item Far from being limited by its implementation size and
         universality, our tool \emph{generates a larger number of valid,
         non-equivalent mutants} for every language we examined than
         any competing, language-specific tool.  That these additional
         mutants
         are meaningful, rather than obscure, is further supported by
         the fact that our mutation scores are similar to those for
         other tools.

@@

In Python regular expression syntax, this expresses the replacement of
strings with a possibly-empty amount of white space followed by a
character pattern indicating a line of source code, with the same
source code, preserving indentation, turned into a comment.
While much more expressive than simple string replacement, regular
expressions are familliar to most developers.
%In principle,
%development of complex regular expressions for operators can even be
%performed by simply providing examples~\cite{bartoli2014automatic}  or
%describing the operator in natural
%language~\cite{zhong2018generating}.  
%These approaches speak to the potential of string manipulation-based approaches
%for mutation testing, without the need for program parsing or analysis.

@@

\begin{table}[hbtp]
\centering
\small 
\caption{Regex vs. Comby for Python. LOC describes the project files
  considered. Muts. is total number of mutants generated for those
  files.  Invalid mutants are ones that fail module load.  We use TCE to check for redundant Python mutants, shown under ``Red.''}
\label{tab:table_python1}
\begin{varwidth}{.9\columnwidth}
\begin{tabularx}{\linewidth}{l|r|rcrc|rcrc}
\toprule
\toprule
                 &                 & \multicolumn{4}{c|}{\textbf{Regex}}             &\multicolumn{4}{c}{\textbf{Comby}}  \\[1ex]
\textbf{Project} & \textbf{LOC}    & \textbf{Muts.} & \textbf{Red.}  & \textbf{\% Valid} & \makecell{\textbf{Mut.} \\ \textbf{Score}} & \textbf{Muts.} & \textbf{Red.}   &  \textbf{\% Valid} & \makecell{\textbf{Mut.} \\ \textbf{Score}}\\\midrule
keract           & 402             & 3984  & 729   & 41.53\% & 0.28           & 3067  & 533    & 47.96\%    & 0.26\\
dtw              & 118             & 2106  & 567   & 46.20\%    & 0.52           & 1835  & 639    & 47.19\%  & 0.55 \\
notion-          & \mr{2}{189}     & \mr{2}{1038}  & \mr{2}{82}    & \mr{2}{18.89\%}   & \mr{2}{0.57}         & \mr{2}{1019}  & \mr{2}{122}    & \mr{2}{19.23\%} & \mr{2}{0.60}  \\
sdk-py           & & & & & & \\[0.5ex]
atlassian-       & \mr{2}{236}      & \mr{2}{1511}  & \mr{2}{207}   & \mr{2}{19.99\%}    & \mr{2}{0.64}        & \mr{2}{1319}  & \mr{2}{351}    & \mr{2}{25.25\%} & \mr{2}{0.62}  \\
python-api       & & & & & & \\[0.5ex]
ESD              & 698             & 6268  & 1642  & 29.46\%    & 0.07           & 3874  & 614    & 44.18\% & 0.07\\
fiber            & 358             & 1565  & 163   & 31.08\% & 0.62           & 1466  & 241    & 31.24\% & 0.64\\\bottomrule
\end{tabularx}
\begin{tabularx}{\linewidth}{Xllrrrrc}
                  && &  \textbf{Red.}     & \textbf{Valid}  & \textbf{Invalid}  & \textbf{\% Valid} &\textbf{Mut. Score}\\\midrule
\multirow{4}{*}{\textbf{Overall}} & \multirow{2}{*}{Regex} & Mean &  565.00 & 913.00 & 1267.33 & 31.19\%  & 0.45\\
    &   &  Std.Dev. & 584.79  & 708.93 & 813.35 & 11.06\%  & 0.23\\\cline{2-8}
 & \multirow{2}{*}{Comby} & Mean & 416.67 & 839.33 & 840.67 & 35.84\% & 0.46\\
  &   &  Std.Dev. &  211.62 & 628.79 & 418.99 & 12.28\%  & 0.24 \\\bottomrule
\end{tabularx}
\end{varwidth}
\end{table}

@@

We do not report the time taken for mutation generation and testing.
Dextools uses mutant schema, and PIT performs bytecode mutation, so
both are much faster for compiling mutants, but other tools took similar
amounts of time to compile mutants.   Mutant execution time was also
generally similar, except for tools that produced very few mutants.
We expect that our approach is generally slower for initial mutant
generation, due to a higher number of invalid mutants, but that this
cost is relatively small compared to the time required to compile and
execute valid mutants for almost all projects.  An informal study of
large Java projects by the author of the PIT gradle
plugin~\cite{mutantSlow}
  suggests that even a successful compile of a typical project takes
  somewhere between one-sixth to one-third the time required for
  executing tests, and failed compiles are usually faster than
  successful ones, since invalid mutants almost always fail during
  parsing, and so code generation, optimization, and linking are not performed.

@@

The ease of defining mutation operators,
plus our interest in generating large numbers of mutants and
prioritizing the most interesting, in order to make mutation testing useful for
projects with extremely high quality test suites, or even using
projects using formal
verification (as in the work of Groce et al.~\cite{groce2018verified}), has encouraged us to take a ``let a thousand flowers bloom''
approach to choosing mutation operators.  For example, our approach
provides many more control-flow mutations than most tools, due to our
observation that test suites sometimes fail to check for loop
execution counts other than one and zero.  Few other tools introduce
{\tt break} and {\tt continue} statements into programs, or transform
one into the other.  These mutants are often equivalent (in fact, they
give rise to a large number of our equivalent mutants reported above), but \emph{also
sometimes expose subtle weaknesses in a high quality test suite}, in our
experience.  In general, we take our philosophy of mutation from the
proposal by Gopinath et al.~\cite{Limits,gopinath2017mutation} that
expanding the set of mutation operations and classes of such
operations, in the long run, is more important than using only a small
set of ``good'' operators, in that the end-goal of mutation testing is
to identify weaknesses in test suites, not simply to estimate a single
score.   The price paid for this
expansiveness in allowing ways to detect weaknessess in testing is a
larger number of mutants to execute, and a slightly larger number of
equivalent mutants.  However, this is not a
fundamental limitation or forced choice.  We propose as future work to
split the operators for each level of the hierarchy into
core operators and a set of ``aggressive'' operators
that includes our full current set of operators.

In fact, this brings us back to a point made in the introduction.
While our approach introduces some unavoidable costs in the
(relatively cheap, compared to execution and analysis) compilation stage of
mutation testing, it may offer \emph{more} flexibility in addressing
execution and analysis stages.  Defining limited custom sets of
operators to run is trivial in our setting, making it easy to
execute only, e.g., statement mutations, or, in a numeric-computation
intensive codebase, only arithmetic operators.  Expanding the
hierarchical structure to include ``famillies'' of mutation operators
is trivial, and gives users great control over mutation.  Our approach
produces source
copies of all mutants, making it easy to apply techniques
like Predictive Mutation Testing~\cite{zhangPMT} that work with source
code.  This also makes it easy to integrate, as we already have done, heuristic
aids to mutant \emph{analysis} such as ranking of unkilled mutants by
similarity and ``interstingness,'' an approach that has already proven
useful in using mutation testing to improve static analysis tools~\cite{StatMut}.
\subsection{Haskell Case Study}

@@

These mutants produce a mutation
score of 0.8 for the two properties defined in the original MuCheck
paper.  In other words, like MuCheck, our tool is able to
determine that the properties provided are insufficient.  In
particular, while they check idempotence and that the result is
sorted, they do not check that the result of {\tt qsort} is a permutaton of the input
list.  The MuCheck paper only provides mutation scores for the two
properties independently.  For idempotence, MuCheck gives a mutation
score of 0.84, and our implementation yields a score of 0.8.  Checking
only sortedness, MuCheck yields a score of 0.61, while our approach
produces an even lower score of 0.4.  Our lower
scores are not due to equivalent mutants: adding a property to check
that the sorted list is a permutation of the original list kills all
our mutants.  Our implementation does not provide the small set of Haskell-specific
mutations provided by MuCheck (e.g., type-aware function replacement).  It nonetheless produces clearly useful
mutants, even for a language arguably radically different from those
for which the universal rules were developed.

@@

%\begin{lstlisting}[language=haskell,caption={\textbf{GPT-4 Generated Haskell Mutation Tool}},label=lst:mutationTool]
%-- Perform mutation on a module
%mutateModule :: Module -> Module

%-- Perform mutation on an expression
%mutateExp :: Exp -> Exp
%mutateExp (If cond trueBranch falseBranch) = If cond falseBranch trueBranch
%mutateExp x = x
%...
%\end{lstlisting}

\begin{lstlisting}[caption={\textbf{Example of GPT-4 Mutants for a Simple Haskell Program}},label=lst:haskellmutants]
[0..] ==> [1..]
IO () ==> IO String  
"Current TODO list:" ==> "Updated TODO list:"
"Invalid command: " ==> "Unrecognized command: "
\end{lstlisting}

@@

To understand how our approach compares with state of the art large language models (LLM), we prompted GPT-4 to generate a mutation testing tool for Haskell.
Surprisingly, the generated program had no syntax errors, and even ran to produce a mutant. However, the program only implemented
the inverting conditionals operation and some basic module mutations,
a small subset of all the operators our tool implements, despite our
tool being unaware of Haskell's existence.. Asking the LLM to expand the set of rules
 produces more operators; however the number of mutants generated by
 such an approach is still a small subset of what our approach is
 capable of doing, even for a language we do not technically support.

 Another approach is to have LLMs generate mutants for a given program. We experiment with prompting GPT-4 to generate mutants for a simple Haskell TODO-list management program.
 \cref{lst:haskellmutants} shows the mutants that GPT-4 generateed for this program (formatted as before ==> after). Notice that some of the mutants generated make sense, and 
 resemble bugs a developer might make (replacing 0 with 1 or using the
 wrong data type). However, GPT-4 also generated mutants changing
strings for all outputs to the console. These are not fruitful
mutants; the messages (potentially) could be checked,
 but the arbitrary nature of such strings makes them
 poor mutants (which is why no mutation tools we are aware of
 implement such an operator).

While LLMs are not able to generate fully working language specific
mutation testing tools from scratch, they are capable of generating
rules in our formats
 that are specific to a given language. For example, prompting GPT-4
 to generate Haskell rules for our tool generates the set of rules seen in \cref{lst:haskellcomby}. The generated rules are 
 both valid Comby syntax and language-specific. E.g., replacing {\tt
   head} with {\tt tail} is a useful Haskell-specific rule.  This suggests that while extending other tools to include new rules
 often involves writing complex code to manipulate ASTs, our tool may
 be extensible simply by requesting an LLM to implement a natural
 language description of the desired rule(s), since GPT-4 already knows
 how to use Comby (and write regular expressions).

@@

Hariri et al. compared C mutation approaches at the source and
compiler IR levels~\cite{CompareSrcBinary} and found that overall
source level mutation was better, producing fewer mutants overall, but
matching the IR approach in the important measures of surface and
minimal mutants and overall mutation score.  Numerous studies compare
Java mutation tools~\cite{MajorPIT,gopinath2017does}, including a
recent article for a more general audience in Communications of the
ACM~\cite{CommACMJavaTool} (perhaps showing the growing interest in
practical mutation testing), which showed that users
considered active maintenance, support for a variety of testing
frameworks, and support for recent Java versions as the most important
features in a Java mutation tool.  The approach proposed in this paper
by its nature tends to promise all three of these key factors without
imposing burdens on maintainers. More recently, there has been some effort to create a 
language agnostic mutation testing tool~\cite{wodeltest}.  However, this tool uses
Wodel, a new (and hard-to-read, compared to Comby) DSL for expressing operators, and requires users to
define a set of operators for each language, without sharing  
of rules.  An
  additional paper on this general topic is a short tool paper for an initial
  prototype of our current tool, omitted for blinding.

@@

As future work, we plan to give users more control over which mutation
operators are used, with controls common to multiple languages.

%We
%add that our approach in general seems to be a good fit to modern
%large software projects, which are often
%multilingual~\cite{Multilingual}: a single mutation tool can be used
%for an entire project.
